Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation

### 2013-03-23 manuscript/manuscript.tex
updated for calls extracted in hadoop

\begin{abstract}
The detection of genomic structural variations remains one of the the most difficult challenges in analyzing high-throughput sequencing data. Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based detection methods. However, the computational requirements for storing and processing data sets with multiple mappings can be formidable. Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computation to cloud or commodity servers. MapReduce is becoming a standard framework for distributing processing across such compute clusters. We describe a novel conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome. In this framework, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome. On simulated and real data sets, our approach achieves performance similar to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes. In particular, our algorithm excels at discovering deletions of size 40bp-100bp in repetitive regions of the genome. In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples. Our implementation and source code are available at \url{http://sonmezsysbio.org/software/cloudbreak}.

@@

Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets. See \textcite{Alkan:2011p547} for a review. Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}. Traditionally, this involves separating mappings into vthose that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment. These approaches then cluster the discordant mappings to find SVs with support from multiple discordantly mapped read pairs. Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome to infer the presence of SVs \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}. Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.

@@

Our formulation divides processing into two distinct MapReduce jobs. The first job finds possible alignment locations for pairs of reads in FASTQ format. The second job computes features from the set of alignment pairs that are relevant to each of a set of small intervals covering the genome. In a final third stage executed outside of MapReduce, the features for each location are exported from the Hadoop file system and post-processed to produce a set of SV calls which span multiple genomic intervals. This MapReduce framework is summarized below with a detailed explanation following.

\algrenewcommand\algorithmicprocedure{\textbf{job}}
  \begin{algorithmic}[1]
    \Procedure{Alignment}{}
    \Function{Map}{$\textrm{ReadPairId }rpid, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
    \ForAll{$ \textrm{Alignments }a \in \textsc{Align}(<s,q>)$}
    \State $\textsc{Emit}(\textrm{ReadPairId }rpid, \textrm{Alignment }a)$
    \EndFor
    \EndFunction
    \Function{Reduce}{$\textrm{ReadPairId }rpid, \textrm{Alignments }a_{1,2,\ldots}$}
    \State $\textrm{AlignmentPairList }ap \gets \textsc{ValidAlignmentPairs}(a_{1,2,\ldots})$
    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList } ap)$
    \EndFunction
    \EndProcedure

    \Procedure{Compute SV Features}{}
    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2>  \in ap$}
    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
    \State $ \textrm{ReadPairInfo }rpi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }rpi)$
    \EndFor
    \EndFor
    \EndFunction
    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
    \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l)$
    \EndFunction
    \EndProcedure
    \State $\textrm{StructuralVariationCalls } svs \gets \textsc{PostProcess}(\phi_{1,2,\ldots})$ 
  \end{algorithmic}

@@

In the second job, we compute a set of features for each location in the genome. To begin, we tile the genome with small fixed-width, non-overlapping intervals. For the experiments reported in this paper we use an interval size of 25bp. Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome. Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads. Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$. For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g. the fragment size implied by the alignments and the likelihood the alignments are correct. We then leave three functions to be implemented depending on the application:
\begin{flalign*}
 \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
 \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
 \textsc{PostProcess} :& \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\} 
\end{flalign*}

The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments. We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair. The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection. Finally, the \textsc{PostProcess} function maps the complete set of features for each locus across the genome into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.

We describe one possible set of functions below; however, we believe that there are many possible definitions that would be useful for different types of SV detection problems that vary in event type sought, depth of coverage of the experiment, heterogeneity of the sample, etc.

@@

We use the following criteria to define a true prediction given a gold standard set of deletion and insertion variants to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been already been discovered by another prediction from the same method. For evaluating insertions, each algorithm produces insertion predictions that define an interval in which the insertion is predicted to have occurred with start and end coordinates $s$ and $e$ as well as the predicted length of the insertion, $l$. The true insertions are defined in terms of their actual insertion coordinate $i$ and their actual length $l_a$. Given this information, we modify the overlap criteria in a) to include overlaps of the intervals $\langle s,\max{\left(e,s+l\right)} \rangle$ and $\langle i,i+l_a \rangle$. Furthermore, we find that Pindel is very accurate at discovering deletions less than or equal to 40bp in length, so we exclude from consideration calls that match a true deletion of less than 40bp where the call is less than or equal to 75bp in length. We also discard all Pindel calls of less than or equal to 20bp. It should be noted that the methods tested here vary in their breakpoint resolution: SR methods have higher resolution than RP methods. Cloudbreak sacrifices additional resolution (by dividing the genome into 25bp windows); we strive however to increase sensitivity and specificity in the hopes that such calls could still be useful even if their resolution is limited. This seems reasonable to us given the possibility of pipelines in which RP calls are refined by SR mappings or local assembly methods.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_DELS_ROC.pdf}
\caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with insertions and deletions from the Venter genome added to one or both haplotypes. Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.}
\label{chr2DeletionsRoc}
\end{figure}

Figure \ref{chr2DeletionsRoc} shows an ROC plot of the performance of each algorithm for detecting deletions on the simulated data set. All approaches show excellent specificity at high thresholds in this simulation. Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY.

Choosing the correct operating point or threshold can be difficult when operating on a new data set, but the use of simulated data and ROC curves allows for some investigation of the perfomance characteristics of algorithms at varying levels. First, we used the results on simulated deletions to characterize the predictions made by each algorithm when a low false discovery rate is required. Table \ref{chr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 10\% based on the ROC curve, representing cutoffs of 2.18 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 12 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and a simple score of 13 Pindel. Cloudbreak shows excellent sensitivity at detecting deletions from each size class, and contributes a number of exclusive true predictions in both small and medium size classes. The output which we obtained from MoDIL did not have a threshold or operating point that correlated with the tradeoff between precision and recall; therefore it is not included in this table.

\begin{table}[t]
\begin{center}
\begin{tabular}{rrrrrr}
  \hline
 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number & 224 &  84 & 82 &  31 & 26\\ 
  \hline
  Cloudbreak  &   50 (   8)  &   50 (\textbf{   3}) &  \textbf{  55} (\textbf{   4}) & \textbf{  12} (\textbf{   4}) & \textbf{  15} (\textbf{0}) \\ 
  Breakdancer &   52 (   9)  &   49 (   2) &    49 (0) &    7 (0) &   14 (\textbf{0}) \\ 
  GASVPro     &   31 (   4)  &   25 (0) &    23 (0) &    2 (0) &    6 (\textbf{0}) \\ 
  DELLY       &   22 (   2)  & \textbf{  56} (\textbf{   3}) &    40 (0) &    8 (0) &   12 (\textbf{0}) \\ 
  Pindel      & \textbf{  60} (\textbf{  34})  &   16 (0) &    41 (   2) &    1 (0) &   12 (\textbf{0})\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2DeletionPredsFDR10}
\end{table}

We also characterized the predictions made by each algorithm at the operating point which gives them maximum sensitivity. For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low. The results are summarized in Table \ref{chr2DeletionPredsMaxSensitivity}. MoDIL has the greatest sensitivity but the precision is extremely low.

\begin{table}[b]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
  \hline
  Cloudbreak    &  \textbf{0.497} & 0.642 & \textbf{ 140} (   5)  &   58 (\textbf{   4}) &    61 (   1) &   13 (0) &   15 (0) \\ 
  Breakdancer   &  0.356 & 0.49 &   89 (0)  &   54 (0) &    53 (0) &    8 (0) &   15 (0) \\ 
  GASVPro        & 0.146 & 0.432 &   83 (   2)  &   32 (0) &    55 (0) &    8 (0) &   15 (0) \\ 
  DELLY           & 0.457 & 0.613 &  114 (   8)  & \textbf{  68} (   2) &  \textbf{  66} (   1) &    9 (0) &   17 (   1) \\ 
  Pindel           & 0.462 & 0.421 &   96 (\textbf{  14})  &   24 (0) &    48 (0) &    5 (0) &   15 (0)\\ 
  MoDIL           & 0.132  & \textbf{0.66} &  123 (   6)  &   66 (   2) &  \textbf{  66} (\textbf{   8}) & \textbf{  17} (\textbf{   6}) & \textbf{  23} (\textbf{   8})\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivy, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2DeletionPredsMaxSensitivity}
\end{table}

\begin{figure}[t]
\centering
\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_INS_ROC.pdf}
\caption{ROC curve of insertion prediction performance on the chromosome 2 simulated data set.}
\label{chr2InsertionsRoc}
\end{figure}

An ROC curve for insertion prediction performance is shown in Figure \ref{chr2InsertionsRoc}. Cloudbreak shows excellent specificity and greater sensitivity than Breakdancer, the other read-pair based method that predicts insertions. The number of insertions identified by each algorithm at maximum sensitivity, again broken down by size, is shown in Table \ref{chr2InsertionPreds}.

\begin{table}[t]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
  \hline
  Cloudbreak   &\textbf{0.624} & 0.194  & \textbf{  65} (  21)  &    9 (   4) &     3 (   1) &    1 (0) & 0 (0) \\ 
  Breakdancer & 0.262 & 0.0968  &   23 (   8)  & \textbf{  14} (\textbf{  10}) &     2 (   1) & 0 (0) & 0 (0) \\ 
  Pindel          & 0.572 & \textbf{0.196} &   52 (\textbf{  25})  &    5 (   3) &  \textbf{  10} (\textbf{  10}) & \textbf{   3} (\textbf{   3}) & \textbf{   9} (\textbf{   9})\\ 
  MoDIL          & 0.186 & 0.0521 &   14 (   1)  &    4 (   1) &     1 (0) &    2 (   1) & 0 (0)\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those insertions that were discovered exclusively by each tool (in parentheses). The total number of insertions in each size class in the true set of insertions is shown in the second row of the header.}
\label{chr2InsertionPreds}
\end{table}

@@

We then tested the predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X. The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those cutoffs, is shown in Table \ref{NA18507DeletionPreds}. Cloudbreak scores well in every category: it identifies more small deletions than any tool except for Pindel, which has low specificity at this cutoff, and contributes more exclusive true positives in larger size classes than any other tool.

>>>>

\begin{abstract}
The detection of genomic structural variations remains one of the the most difficult challenges in analyzing high-throughput sequencing data. Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based detection methods. However, the computational requirements for storing and processing data sets with multiple mappings can be formidable. Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computation to cloud or commodity servers. MapReduce is becoming a standard framework for distributing processing across such compute clusters. We describe a novel conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome. In this framework, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome. On simulated and real data sets, our approach achieves performance similar to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes. In particular, our algorithm excels at discovering insertions and deletions of size 40bp-100bp, and those in repetitive regions of the genome. In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples. Our implementation and source code are available at \url{http://sonmezsysbio.org/software/cloudbreak}.

@@

Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets. See \textcite{Alkan:2011p547} for a review. Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}. Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment. These approaches then cluster the discordant mappings to find SVs with support from multiple discordantly mapped read pairs. Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome to infer the presence of SVs \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}. Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.

@@

Our formulation divides processing into three distinct MapReduce jobs. The first job finds possible alignment locations for pairs of reads in FASTQ format. The second job computes features from the set of alignment pairs that are relevant to each of a set of small intervals covering the genome. In the third stage, the features for all locations on each chromosome are examined to find contiguous blocks of locations that support an SV call. This MapReduce framework is summarized in Algorithm 1.

\begin{algorithm}
\algrenewcommand\algorithmicprocedure{\textbf{job}}
  \begin{algorithmic}[1]
    \Procedure{Alignment}{}
    \Function{Map}{$\textrm{ReadPairId }rpid, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
    \ForAll{$ \textrm{Alignments }a \in \textsc{Align}(<s,q>)$}
    \State $\textsc{Emit}(\textrm{ReadPairId }rpid, \textrm{Alignment }a)$
    \EndFor
    \EndFunction
    \Function{Reduce}{$\textrm{ReadPairId }rpid, \textrm{Alignments }a_{1,2,\ldots}$}
    \State $\textrm{AlignmentPairList }ap \gets \textsc{ValidAlignmentPairs}(a_{1,2,\ldots})$
    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList } ap)$
    \EndFunction
    \EndProcedure

    \Procedure{Compute SV Features}{}
    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2>  \in ap$}
    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
    \State $ \textrm{ReadPairInfo }rpi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }rpi)$
    \EndFor
    \EndFor
    \EndFunction
    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
    \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l)$
    \EndFunction
    \EndProcedure

    \Procedure{Call SVs}{}
    \Function{Map}{$\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l$}
    \State $\textsc{Emit}(\textrm{Chromosome}(l), <l,\phi_l>)$
    \EndFunction
    \Function{Reduce}{$\textrm{Chromosome }c, \textrm{GenomicLocation } l_{1,2,\ldots},\phi_{1,2,\ldots}$}
    \State $\textrm{StructuralVariationCalls } svs_c \gets \textsc{PostProcess }(\phi_{1,2,\ldots})$
    \EndFunction
    \EndProcedure
  \end{algorithmic}
\label{cloudbreakAlgorithm}
\caption{The algorithmic framwork for SV calling in MapReduce.}
\end{algorithm}

@@

In the second job, we compute a set of features for each location in the genome. To begin, we tile the genome with small fixed-width, non-overlapping intervals. For the experiments reported in this paper we use an interval size of 25bp. Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome. Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads. Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$. For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g. the fragment size implied by the alignments and the likelihood the alignments are correct. We then leave two functions to be implemented depending on the application:
\begin{flalign*}
 \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
 \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
\end{flalign*}

The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments. We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair. The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.

Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location. It calls another application-specific function  $\textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\}$  that maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$. We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task.

We describe our implementations of the $\textsc{Loci}$, $\Phi$, and $\textsc{PostProcess}$ functions below. However, we believe that there are many possible definitions that would be useful for different types of SV detection problems that vary in event type sought, depth of coverage of the experiment, heterogeneity of the sample, etc.

@@

We use the following criteria to define a true prediction given a gold standard set of deletion and insertion variants to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been already been discovered by another prediction from the same method. For evaluating insertions, each algorithm produces insertion predictions that define an interval in which the insertion is predicted to have occurred with start and end coordinates $s$ and $e$ as well as the predicted length of the insertion, $l$. The true insertions are defined in terms of their actual insertion coordinate $i$ and their actual length $l_a$. Given this information, we modify the overlap criteria in a) to include overlaps of the intervals $\langle s,\max{\left(e,s+l\right)} \rangle$ and $\langle i,i+l_a \rangle$. In this study we are interested in detecting events larger than 40bp, because with longer reads, smaller events can be more easily discovered by examining gaps in individual reads. Both Pindel and MoDIL make many calls with a predicted event size of under 40bp, so we remove those calls from the output sets of those programs. Finally, we exclude from consideration calls from all approaches that match a true deletion of less than 40bp where the predicted variant length is less than or equal to 75bp in length.

It should be noted that the methods tested here vary in their breakpoint resolution: SR methods have higher resolution than RP methods. Cloudbreak sacrifices additional resolution (by dividing the genome into 25bp windows); we strive however to increase sensitivity and specificity in the hopes that such calls could still be useful even if their resolution is limited. This seems reasonable to us given the possibility of pipelines in which RP calls are refined by SR mappings or local assembly methods.

@@

\begin{figure}[b]
\centering
\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_DELS_ROC.pdf}
\caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with insertions and deletions from the Venter genome added to one or both haplotypes. Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.}
\label{chr2DeletionsRoc}
\end{figure}

Figure \ref{chr2DeletionsRoc} shows an ROC plot of the performance of each algorithm for detecting deletions on the simulated data set. All approaches show excellent specificity at high thresholds in this simulation. Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY. The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the tradeoff between precision and recall; therefore it is not included in this figure.

Choosing the correct operating point or threshold can be difficult when operating on a new data set, but the use of simulated data and ROC curves allows for some investigation of the perfomance characteristics of algorithms at varying levels. First, we used the results on simulated deletions to characterize the predictions made by each algorithm when a low false discovery rate is required. Table \ref{chr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve, representing cutoffs of 2.22 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 11 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and a simple score of 13 for Pindel. Cloudbreak shows good performance at detecting deletions in all size classes, and finds more medium and large deletions than the other approaches. Pindel exlusively identifies a large number of small (40bp-100bp) deletions, as is expected for a split-read method.

\begin{table}[b]
\begin{center}
\begin{tabular}{rrrrrr}
  \hline
 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number & 224 &  84 & 82 &  31 & 26\\ 
  \hline
  Cloudbreak  &   48 (   8)  &   49 (\textbf{   2}) &  \textbf{  55} (\textbf{   4}) & \textbf{  12} (\textbf{   4}) & \textbf{  15} (\textbf{0}) \\ 
  Breakdancer &   52 (  10)  &   49 (\textbf{   2}) &    49 (0) &    7 (0) &   14 (\textbf{0}) \\ 
  GASVPro     &   35 (   5)  &   26 (0) &    26 (0) &    2 (0) &    6 (\textbf{0}) \\ 
  DELLY       &   22 (   2)  & \textbf{  56} (\textbf{   2}) &    40 (0) &    8 (0) &   12 (\textbf{0}) \\ 
  Pindel      & \textbf{  60} (\textbf{  34})  &   16 (0) &    41 (   2) &    1 (0) &   12 (\textbf{0})\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2DeletionPredsFDR10}
\end{table}

We also characterized the predictions made by each algorithm at the operating point which gives them maximum sensitivity. For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low. The results are summarized in Table \ref{chr2DeletionPredsMaxSensitivity}. MoDIL has the greatest recall but low precision. Cloudbreak has high precision and recall at this threshold, and surprisingly discovers more small deletions than Pindel.

\begin{table}[t]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
  \hline
  Cloudbreak    &  \textbf{0.507} & 0.649 & \textbf{ 144} (   6)  &   57 (\textbf{   4}) &    61 (   1) &   13 (0) &   15 (0) \\ 
  Breakdancer   &  0.356 & 0.49 &   89 (0)  &   54 (0) &    53 (0) &    8 (0) &   15 (0) \\ 
  GASVPro        & 0.146 & 0.432 &   83 (   2)  &   32 (0) &    55 (0) &    8 (0) &   15 (0) \\ 
  DELLY           & 0.457 & 0.613 &  114 (   8)  & \textbf{  68} (   2) &  \textbf{  66} (   1) &    9 (0) &   17 (   1) \\ 
  Pindel           & 0.462 & 0.421 &   96 (\textbf{  13})  &   24 (0) &    48 (0) &    5 (0) &   15 (0)\\ 
  MoDIL           & 0.132  & \textbf{0.66} &  123 (   6)  &   66 (   2) &  \textbf{  66} (\textbf{   8}) & \textbf{  17} (\textbf{   6}) & \textbf{  23} (\textbf{   8})\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivy, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2DeletionPredsMaxSensitivity}
\end{table}

\begin{figure}[b]
\centering
\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_INS_ROC.pdf}
\caption{ROC curve of insertion prediction performance on the chromosome 2 simulated data set.}
\label{chr2InsertionsRoc}
\end{figure}

An ROC curve for insertion prediction performance is shown in Figure \ref{chr2InsertionsRoc}. The number of insertions identified by each algorithm at maximum sensitivity, again broken down by size, is shown in Table \ref{chr2InsertionPreds}. Cloudbreak shows excellent precision, although recall is low for all four aproaches. Cloudbreak again identifies the most small variants. Pindel is the only tool which can identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by read-pair mapping.

\begin{table}[t]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
  \hline
  Cloudbreak   &\textbf{0.769} & \textbf{0.199}  & \textbf{  67} (  24)  &    9 (   3) &     3 (   1) &    1 (0) & 0 (0) \\ 
  Breakdancer & 0.262 & 0.0968  &   23 (   8)  & \textbf{  14} (\textbf{   9}) &     2 (   1) & 0 (0) & 0 (0) \\ 
  Pindel          & 0.572 & 0.196 &   52 (\textbf{  26})  &    5 (   3) &  \textbf{  10} (\textbf{  10}) & \textbf{   3} (\textbf{   3}) & \textbf{   9} (\textbf{   9})\\ 
  MoDIL          & 0.186 & 0.0521 &   14 (   1)  &    4 (   1) &     1 (0) &    2 (   1) & 0 (0)\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those insertions that were discovered exclusively by each tool (in parentheses). The total number of insertions in each size class in the true set of insertions is shown in the second row of the header.}
\label{chr2InsertionPreds}
\end{table}

@@

We then tested the predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X. The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those thresholds, is shown in Table \ref{NA18507DeletionPreds}. Cloudbreak has the greatest sensitivity of any tool at these thresholds, identifying the most variants in each size class.
